Negative Selection in the 20th Century

### 2020-05-22 why-negative-selection.Rmd
Work on reporting results of weighting

* If this theory is true then the pattern of differences is explained by
  (potential) earnings. So the strength of the relationship of a PGS with 
  number of children should be explained by the strength of its relationship
  with earnings and education.
* Even after weighting, effect sizes for siblings are still higher than for
  children. Why?
  - Remaining ascertainment bias?
  - Siblings don't include those with 0 children?
  - Something else?
* If we weight with the GHS, and plot the means of figure 1 using this w
  weighting, most things stay the same - but the change in EA is reversed! EA
  increases over time. Could this be because the sample includes many
  relatively educated older people? If so, then mean effects are *overestimated*
  by figure 1.
* Could be useful: f.2139 - age first had sex (includes "never had sex" 
  which may explain some of the many NAs for f.2141, num sex partners)
*  Need f.6138 to correctly calculate `age_fulltime_edu` including
  those who went to college.
* Look at geography, esp. in siblings regressions.
* Work more on weighting the data
  - check for existing work by others?
  - deal with the fact that GHS 06 has its own weights and use
    subset.svydesign to reweight?
* control for age in n_partners regressions

@@

res_w_wide <- full_join(
        res_weighted,
        res_sibs_parent_weights,
        by = "term",
        suffix = c(".children", ".sibs")
      ) 
# barbaric but it works:
res_uw_wide <- left_join(
  res_all %>% filter(dep.var == "N children", reg.type == "raw"),
  res_all %>% filter(dep.var == "N siblings", reg.type == "raw"),
  by = "term",
  suffix = c(".children", ".sibs")
)

res_wide <- left_join(res_w_wide, res_uw_wide, by = "term", 
      suffix = c(".wtd", ".uwtd"))

res_wide %<>% mutate(
        prop_children = estimate.children.wtd/estimate.children.uwtd,
        prop_sibs     = estimate.sibs.wtd/estimate.sibs.uwtd
      )

@@

We rerun the basic regressions for both children and siblings, using the new
weights. Figure \@ref(fig:plot-res-weighted) shows the results. The weightings
are fairly successful at reconciling the overall size of effects from
siblings and children regressions. Tables \@ref(tab:tbl-sibs-weights) and
\@ref(tab:tbl-children-weights) show the largest and smallest changes due to 
weighting in siblings and children regressions.

```{r tbl-sibs-weights}

stars <- function (x) ifelse(x < 0.05/33, "*", "")

top_bottom_5 <- function (x, caption) {
  x %>% 
      slice(c(1:5, nrow(.) - 0:4)) %>% 
      as_hux() %>% 
      insert_row(rep("...", 4), after = 6) %>% 
      set_number_format(-1, -1, 3) %>% 
      set_number_format(-1, 4, 0) %>% 
      set_align(-1, -1, ".") %>% 
      theme_basic() %>% 
      set_font_size(8) %>% 
      set_caption(caption) %>% 
      set_caption_pos("bottom") %>% 
      set_caption_width(1) %>% 
      add_footnote("Insignificant effects not shown", font_size = 7) %>% 
      set_all_padding(0) 
}

res_wide %>% 
      arrange(desc(prop_sibs)) %>% 
      select(term, matches("estimate.sibs"), matches("p.value.sibs"), 
        prop_sibs) %>% 
      filter(p.value.sibs.uwtd < 0.05/33) %>% 
      transmute(
        Score = pretty_names(term),
        `Weighted estimate` = estimate.sibs.wtd,
        `Unweighted estimate` = estimate.sibs.uwtd,
        `Proportion (%)` = prop_sibs * 100
      ) %>% 
      top_bottom_5(caption = "Effects of weighting on sibling regressions: top and bottom 5 changes")
```

```{r tbl-children-weights}

res_wide %>% 
      arrange(desc(prop_children)) %>% 
      select(term, matches("estimate.children"), matches("p.value.children"), 
        prop_children) %>% 
      filter(p.value.children.uwtd < 0.05/33) %>% 
      transmute(
        Score = pretty_names(term),
        `Weighted estimate` = estimate.children.wtd,
        `Unweighted estimate` = estimate.children.uwtd,
        `Proportion (%)` = prop_children * 100
      ) %>% 
      top_bottom_5(caption = "Effects of weighting on children regressions: top and bottom 5 changes")
```

```{r plot-res-weighted, fig.cap = "Effects of polygenic scores on number of children/siblings. Weighted regressions."}

@@

We also weighted data for women with children only, by age, education and age 
at first live birth.  Figure \@ref(fig:plot-flb-weights) 
shows the results. Weighting increased effect sizes on average by 
`r 100 * (prop_wt_nowt - 1)` per cent. There 
are large changes for some variables, including EA3 
(`r pct_change_score("EA3_excl_23andMe_UK")` per cent), cognitive
ability (`r pct_change_score("cognitive_ability")` per cent) and
waist-hip ratio (WHR; `r pct_change_score("whr_combined")` percent).

@@

bind_rows(
        Weighted   = res_flb_weights, 
        Unweighted = res_sex %>% filter(sex == "Female") %>% select(-subset, -sex),
        .id        = "Regression"
      ) %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, estimate, order_abs(1))
      ) %>% 
      ggplot(aes(estimate, term, shape = Regression, 
        color = p.value < 0.05/66)) +
        my_vline +
        geom_point(size = 1.5, stroke = 0.7, fill = "steelblue") +
        scale_shape_manual(values = c("Weighted" = 22, "Unweighted" = 21)) +  
        pval_scale() +
        my_labs() + 
        my_theme()
```

These results suggest that previous work may have underestimated the level of
selection occurring in the population. [TODO: who?]

@@

In this section, we dig into the mechanisms causing natural selection in the
population. In particular, why are selection effects *oppositely* signed in
certain subgroups -- for example, in older and younger mothers (by age at first
live birth) or in those with more and fewer lifetime sexual partners?

A possible answer is given by the economic theory of fertility [TODO cite Becker 1960].
According to this, increases in potential earned income affect fertility via two opposing channels. There is an "income effect" by which children become more
affordable, like any other good. Since childrearing has a cost in time, there is
also a substitution effect: the opportunity cost of childrearing increases if
one's market wage is higher. The income effect would lead higher earners to
have more children. The substitution effect would lead higher earners to have
fewer children.

Traditionally, it was assumed that since women
spent more of their time on childcare, the substitution effect would dominate
for women's earnings, while the income effect would dominate for men's earnings. 
However, a more relevant split may be between those raising children alone, and
those raising children in a two-parent household. The substitution effect is likely
to be particularly strong for lone parents, since they have less opportunity
to share childcare responsibilities.

This explanation predicts that the strength of polygenic scores' effects on
fertility will correlate with the strength of their relationship with earnings.
We measure the correlation of each PGS with household income, as well as with
education level, a phenotypic variable which is likely to predict earnings. We
then correlate these measurements with effect sizes on number of children.
Across scores, there is a clear negative correlation between effect on
fertility, and link to both income and education.

Figure \@ref(fig:plot-age-flb) shows the results of *children* regressions for 
women only,
controlling for age at first live birth. Effect sizes are greatly reduced. Partly,
this is because the regressions exclude childless women. However,
in `r opp_signed` out of `r length(score_names)` cases, effect sizes actually
change sign with the controls. The correlation between effect sizes
controlling for age at first live birth, and raw effect sizes, is `r cor_raw_flb`.
```{r plot-age-flb, fig.cap = "Effect sizes on number of children, controlling for age at first live birth (women only). Effect sizes for women without controls are shown for comparison"}

@@

loadd(res_together)
n_regs <- as.double(nrow(res_together))
res_together %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, estimate, order_abs(2)),
        dep.var = dplyr::recode(dep.var, 
          "n_children" = "N Children", "n_sibs" = "N Siblings"),
        dep.var = fct_relevel(dep.var, "N Siblings")
      ) %>% 
      ggplot(aes(y = term, x = estimate, color = p.value < 0.05/{{n_regs}}, 
        fill  = PCs)) + 
        pval_geom_point(alpha = 0.8) +
        my_vline +
        facet_wrap(vars(dep.var)) +
        pval_scale() +
        my_fill_scale(n = 2) +
        my_labs() +
        theme(legend.position = "bottom")
```

>>>>

* If we weight with the GHS, and plot the means of figure 1 using this w
  weighting, most things stay the same - but the change in EA is reversed! EA
  increases over time. Could this be because the sample includes many
  relatively educated older people? If so, then mean effects are *overestimated*
  by figure 1. 
* Need to make the weighting very watertight. 
  - Include income? 
  - Interact age and other variables?
  - Investigate "MRP" (Gelman 2005)
  - Understand reweighting on subsets of survey data.
  - Check for existing work
* Reweight other regressions (e.g. the omnibus regression in the appendix)
* Could be useful: f.2139 - age first had sex (includes "never had sex" 
  which may explain some of the many NAs for f.2141, num sex partners)
*  Need f.6138 to correctly calculate `age_fulltime_edu` including
  those who went to college.
* Look at geography, esp. in siblings regressions.
* control for age in n_partners regressions.

@@

res_wide <- full_join(
        res_weighted,
        res_sibs_parent_weights,
        by = "term",
        suffix = c(".children", ".sibs")
      ) %>% 
      filter(p.value.children < 0.05/33 | p.value.sibs < 0.05/33)

chn_larger <- with(res_wide, estimate.children > estimate.sibs)

@@

We rerun the basic regressions for both children and siblings, using the new
weights. Figure \@ref(fig:plot-res-weighted) shows the results. The weightings
are fairly successful at reconciling the overall size of effects from siblings
and children regressions. Among weighted scores that are significant (at p <
0.05/33) in either regression, `r sum(chn_larger)` estimates are
higher for children and `r length(chn_larger) - sum(chn_larger)` are
lower. However, some estimates differ widely or change
sign across the generations.

```{r plot-res-weighted, fig.cap = "Effects of polygenic scores on number of children/siblings. Regressions weighted by sex, age and education."}

@@

We also weighted data for women with children only, by age, education and age 
at first live birth.  Figure \@ref(fig:plot-flb-weights) 
shows the results. There 
are large changes for some variables, including EA3 
(`r pct_change_score("EA3_excl_23andMe_UK")` per cent), cognitive
ability (`r pct_change_score("cognitive_ability")` per cent) and
waist-hip ratio (WHR; `r pct_change_score("whr_combined")` percent).

Overall, these results suggest that previous work may have underestimated the level of
selection occurring in the population. [TODO: who?]

@@

bind_rows(
        Weighted   = res_flb_weights, 
        Unweighted = res_sex %>% filter(sex == "Female") %>% select(-subset, -sex),
        .id        = "Regression"
      ) %>% 
      standard_ggplot(Regression, n_regs = 66)
```

Why are selection effects *oppositely* signed in certain subgroups -- for
example, in older and younger mothers (by age at first live birth) or in those
with more and fewer lifetime sexual partners? A possible answer is given by the
economic theory of fertility [TODO cite Becker 1960]. According to this,
increases in potential earned income affect fertility via two opposing channels.
There is an "income effect" by which children become more affordable, like any
other good. Since childrearing has a cost in time, there is also a substitution
effect: the opportunity cost of childrearing increases if one's market wage is
higher. The income effect would lead higher earners to have more children. The
substitution effect would lead higher earners to have fewer children. It is
often assumed that the substitution effect will dominate for lone parents, since
they have less opportunity to share childcare responsibilities, while the income
effect will dominate for couples who are able to reap the gains from
specialization.

This explanation predicts that the strength of polygenic scores' effects on
fertility will correlate with the strength of their relationship with earnings.
We measure the correlation of each PGS with household income, as well as with
education level, a phenotypic variable which is likely to predict earnings. We
then relate these measurements to effect sizes on number of children.
Across scores, there is a clear negative relationship between effect on
fertility, and the correlation with income and/or education (Figure \@ref(fig:plot-cors-income-educ).

@@

Figure \@ref(fig:plot-age-flb) shows the results of *children* regressions for 
women only,
controlling for age at first live birth. Effect sizes are greatly reduced. Partly,
this is because the regressions exclude childless women. However,
in `r opp_signed` out of `r length(score_names)` cases, effect sizes actually
change sign with the controls. The correlation between effect sizes
controlling for age at first live birth, and raw effect sizes, is `r cor_raw_flb`.

```{r plot-age-flb, fig.cap = "Effect sizes on number of children, controlling for age at first live birth (women only). Effect sizes for women without controls are shown for comparison"}

@@

loadd(res_together)
n_regs <- as.double(nrow(res_together))
res_together %>% 
      mutate(
        term = pretty_names(term),
        dep.var = dplyr::recode(dep.var, 
          "n_children" = "N Children", "n_sibs" = "N Siblings"),
        dep.var = fct_relevel(dep.var, "N Siblings")
      ) %>% 
      ggplot(aes(y = reorder_within(term, estimate, dep.var, order_abs(1)), 
        x = estimate, color = p.value < 0.05/{{n_regs}}, fill  = PCs)) + 
        pval_geom_point(alpha = 0.8) +
        my_vline +
        facet_wrap(vars(dep.var), scales = "free_y") +
        pval_scale() +
        my_fill_scale(n = 2) +
        my_labs() + 
        scale_y_reordered() +
        theme(legend.position = "bottom")
```
